Stage-Splitting a Modal Language

### 2014-06-19 paper/semantics.tex
Forgot to commit in previous checkin.
put the right value in
adding some code formatting

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
\footnote{We render $\fut$ as {\tt \$} in code examples.}

@@

To achieve this behavior, we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
[1+2/x1]
let x = (next{x1}, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
[1+2/x1]
next{
	prev{#1 (next{x1}, 7)} *
	prev{#1 (next{x1}, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[1+2/x1]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|next {x1 * x1 * 7}|.
\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|1+2|.
\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|1215|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
and a partial value $v$ which may refer to those variables. This judgment
necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
residual.

>>>>

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
%\footnote{We render $\fut$ as {\tt \$} in code examples.}

@@

To achieve this behavior, we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
[x1|->1+2]
let x = (next{x1}, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
[1+2/x1]
next{
	prev{#1 (next{x1}, 7)} *
	prev{#1 (next{x1}, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[x1|->1+2]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|next {x1 * x1 * 7}|.
\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|1+2|.
\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
and a partial value $v$ which may refer to those variables. This judgment
necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
residual.
